A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-28 text/discussion.Rmd
fix textual figure references; reflow text
Update text in manuscript and file names

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes include genome
stability and immortality, as "enabling characteristics" of the cancer phenotype
[@hallmarks_II]. Such pathways are disrupted in most clusters and are frequently
the result of aberrations involving common driver genes such as BRCA1/2, MMR
genes, mitotic checkpoints, cohesion complexes, _etc._ Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful (_e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness. Importantly, our observations do not imply the order in which these
mutations should accumulate. This could be addressed in a future study by
evolutionary analysis of clonality, drawing inference from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to surmise that these mutations promote or enable acquisition of
cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as APOBEC,
BRCA1 and BRCA2 produce mutational hotspots reflecting varying selective
pressures in different tissues [@PCAWG-MAIN].

@@

We report that enrichment in metastatic tumors across all clusters yielded
generally lower effect sizes and larger p-values than the cluster specific
analysis, suggesting that signal is diluted when clusters are pooled, and
supporting the view that metastasis has cluster-specific requirements. Since the
number of metastatic samples is relatively low, this part of our analysis is
likely underpowered and subject to expanded analysis with larger cohorts. The
fact that most metastatic enrichment is cluster-specific and has a tendency to
overlap with cluster-specific pathways from non-metastatic tumors of neighboring
clusters suggests that newly acquired mutations result in similar clusters
converging on one or more deadly phenotypes with critical features of end-stage
cancer. Thus, even at relatively low power, our analysis of metastasis uncovered
differences between tumors against the noisy backdrop of tissue specific
profiles. As a major caveat to our pan-cancer analysis, we acknowledge that many
therapies _do_ target highly-specific driver genes, markers, and signaling
pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader context of
the genetic background and pathway vulnerability of tumors containing such
markers may aid in creating smarter combination therapies. We submit that when
we discover the requirements of each cluster with respect to pathway disruptions
and metastasis we may be able to target them therapeutically and prevent further
adaptation.

>>>>

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes include genome
stability and immortality, as "enabling characteristics" of the cancer phenotype
[@hallmarks_II]. Such pathways are disrupted in most clusters and are frequently
the result of aberrations involving common driver genes such as BRCA1/2, MMR
genes, mitotic checkpoints, cohesion complexes, _etc._ Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful (_e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness. Importantly, our observations do not imply the order in which these
mutations should accumulate. This could be addressed in a future study by
evolutionary analysis of clonality, drawing inference from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to surmise that these mutations promote or enable acquisition of
cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as _APOBEC_,
_BRCA1_ and _BRCA2_ produce mutational hotspots reflecting varying selective
pressures in different tissues [@PCAWG-MAIN].

@@

We report that enrichment in metastatic tumors across all clusters yielded
generally lower effect sizes and larger p-values than the cluster specific
analysis, suggesting that signal is diluted when clusters are pooled, and
supporting the view that metastasis has cluster-specific requirements. Since the
number of metastatic samples is relatively low, this part of our analysis is
likely underpowered and subject to expanded analysis with larger cohorts. The
fact that most metastatic enrichment is cluster-specific and has a tendency to
overlap with cluster-specific pathways from non-metastatic tumors of neighboring
clusters suggests that newly acquired mutations result in similar clusters
converging on one or more deadly phenotypes with critical features of end-stage
cancer. Thus, even at relatively low power, our analysis of metastasis uncovered
differences between tumors against the noisy backdrop of tissue specific
profiles. As a major caveat to our pan-cancer analysis, we acknowledge that many
therapies _do_ target highly-specific driver genes, markers, and signaling
pathways ( _e.g. TP53_, _EGFR_ or _HER2_), but understanding the broader context of
the genetic background and pathway vulnerability of tumors containing such
markers may aid in creating smarter combination therapies. We submit that when
we discover the requirements of each cluster with respect to pathway disruptions
and metastasis we may be able to target them therapeutically and prevent further
adaptation.

### 2020-08-28 text/manuscript.Rmd
Update text in manuscript and file names

To avoid bias toward larger pathways with more genes, we counted them as
disrupted if one or more member genes were mutated. We do not attempt to
calculate enrichment for mutations within a pathway. Data binarized by pathway
are likely to be noisy for several reasons. First, point mutations can be
deleterious (attenuating, hypomorphic or antimorphic) or activating (neomorphic
or hypermorphic) in genes, and these can in turn act functionally as either
oncogenes or tumor suppressors. For this study we assume a significant fraction
of these mutations are generically disruptive to normal pathway activity since
it is impossible to know the tumor promoting effects of all mutations, including
rarely studied genes. Second, we know that low- and non-expressed genes
accumulate mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified low expressed genes in each type of cancer and eliminated
them for that cancer type only (see methods for details). Highly expressed genes
could also have high mutation rates owing to transcription induced mutagenesis
[@parkGenomicEvidenceElevated2012]. This phenomenon would likely result in
cell-type-specific biases that might be biologically meaningful for
predisposition to different classes of cancer however, and therefore chose not
to exclude these genes from our analysis. After selecting our pathways and
genes, we then compiled a matrix of the pathways, assigning a Boolean value of 1
to each pathway with one or more genes mutated and 0 for all others (**figure
1**).

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 1
```{r child='../code/figure1v2.Rmd'}
```

@@

#### Figure 2
```{r child = '../code/figure3.Rmd'}
```

@@

Many cancers have molecular or histological subtypes defined based on gene
expression, pathology or other -omics profiles. These subtypes often have
different standards of care owing to different drug sensitivity (or other
factors). If the histological subtypes represent true molecular phenotypes, one
predicts that they should segregate with our pathway-based clusters, therefore
providing support for the clusters as proxies for molecular pathology
sub-typing. To our surprise, we find a similar result to the previous analysis
of cancer types projected onto the UMAP of pathway disruptions. To illustrate
this, we projected annotations for each of the breast cancer subtypes, composed
of Triple-negative/Basal-like, Her2 positive, normal-like, and luminal A and B
subtypes onto the UMAP. These are among the most heavily studied molecular
subtypes in cancer, each with different prognoses and standards of care. We did
not observe any exclusive segregation by pathway for these subtypes (**Figure
3B**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** and online [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the
subtypes, though often biased to one or more clusters, are almost never
exclusive. We interpret these data in aggregate to mean that the clusters do not
correspond to previously identified molecular subtypes within the parent cancer
types.

@@

#### Figure 3
```{r child = '../code/figure5.Rmd'}
```

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from their tissue-specific -omics data
signatures. Given our pathway-disruption based clustering, this raises the
question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (_i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from
[**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations in genes
responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
subunit genes involved in protein degradation. We also observed that genes that
were enriched for one cluster are depleted from others (_i.e._ is enriched in
cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
depleted in clusters 3 and 8). Next, we investigated the proportion of samples
per cancer type for the significant genes within a cluster (**Figure 4**;
heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
more highly mutated genes across all cancers. Instead, when observing the
mutation rate for these genes within samples that belong to a cluster, the
mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
there is no global pattern, indicating that our clusters are not driven by
individual genes, but rather networks as a whole. Taken together, our data
identify a framework of cancer type-specific mutations associated with specific
clusters.

@@

#### Figure 4
```{r child = '../code/figure6.Rmd'}
```

>>>>

To avoid bias toward larger pathways with more genes, we counted them as
disrupted if one or more member genes were mutated. We do not attempt to
calculate enrichment for mutations within a pathway. Data binarized by pathway
are likely to be noisy for several reasons. First, point mutations can be
deleterious (attenuating, hypomorphic or antimorphic) or activating (neomorphic
or hypermorphic) in genes, and these can in turn act functionally as either
oncogenes or tumor suppressors. For this study we assume a significant fraction
of these mutations are generically disruptive to normal pathway activity since
it is impossible to know the tumor promoting effects of all mutations, including
rarely studied genes. Second, we know that low- and non-expressed genes
accumulate mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified low expressed genes in each type of cancer and eliminated
them for that cancer type only (see methods for details). Highly expressed genes
could also have high mutation rates owing to transcription induced mutagenesis
[@parkGenomicEvidenceElevated2012]. This phenomenon would likely result in
cell-type-specific biases that might be biologically meaningful for
predisposition to different classes of cancer however, and therefore chose not
to exclude these genes from our analysis. After selecting our pathways and
genes, we then compiled a matrix of the pathways, assigning a Boolean value of 1
to each pathway with one or more genes mutated and 0 for all others (**figure
1A**).

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 1
```{r child='../code/figure1.Rmd'}
```

@@

#### Figure 2
```{r child = '../code/figure2.Rmd'}
```

@@

Many cancers have molecular or histological subtypes defined based on gene
expression, pathology or other -omics profiles. These subtypes often have
different standards of care owing to different drug sensitivity (or other
factors). If the histological subtypes represent true molecular phenotypes, one
predicts that they should segregate with our pathway-based clusters, therefore
providing support for the clusters as proxies for molecular pathology
sub-typing. To our surprise, we find a similar result to the previous analysis
of cancer types projected onto the UMAP of pathway disruptions. To illustrate
this, we projected annotations for each of the breast cancer subtypes, composed
of Triple-negative/Basal-like, Her2 positive, normal-like, and luminal A and B
subtypes onto the UMAP. These are among the most heavily studied molecular
subtypes in cancer, each with different prognoses and standards of care. We did
not observe any exclusive segregation by pathway for these subtypes (**figure
2B**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** and online [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the
subtypes, though often biased to one or more clusters, are almost never
exclusive. We interpret these data in aggregate to mean that the clusters do not
correspond to previously identified molecular subtypes within the parent cancer
types.

@@

#### Figure 3
```{r child = '../code/figure3.Rmd'}
```

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from their tissue-specific -omics data
signatures. Given our pathway-disruption based clustering, this raises the
question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (_i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**figure 4**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from
[**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations in genes
responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
subunit genes involved in protein degradation. We also observed that genes that
were enriched for one cluster are depleted from others (_i.e._ is enriched in
cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
depleted in clusters 3 and 8). Next, we investigated the proportion of samples
per cancer type for the significant genes within a cluster (**figure 4**;
heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
more highly mutated genes across all cancers. Instead, when observing the
mutation rate for these genes within samples that belong to a cluster, the
mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
there is no global pattern, indicating that our clusters are not driven by
individual genes, but rather networks as a whole. Taken together, our data
identify a framework of cancer type-specific mutations associated with specific
clusters.

@@

#### Figure 4
```{r child = '../code/figure4.Rmd'}
```

### 2020-08-28 text/survival.Rmd
Update text in manuscript and file names

To illustrate the model, in **Figure 5A** we simulate survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

#### Figure 5
```{r child='../code/survival/survfig.Rmd'}
```

@@

We found that cancer types, as expected, have a range of prognoses relative to
the general population. In **Figure 5B-E** we see three particularly deadly
cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
where diagnosis apparently decreases risk relative to background (Thyroid:
THCA). Cancers with posterior probability for relative risk of less than 1 such
as THCA should be interpreted carefully. This Bayesian model is a model for a
state of information. The information that a person is diagnosed with cancer may
lead us to expect that they will live a shorter time than the general population
of matched age (k > 1), or a longer time than the general population of matched
age (k < 1). A shorter prognosis could result from the cancer being aggressive,
injuring the body and causing death. By contrast, a longer prognosis may result
from the cancer being relatively mild, and therefore diagnosis could be an
informational signal that the patient is health conscious, with the comparison
group having more people whose cancers go undiagnosed. It is important to note
therefore that the diagnosis can increase our expectation of life relative to
the comparison group, even if it decreases the expectation of life of the
individual relative to the counterfactual where they did not have cancer.

The degree to which a given cancer accelerates aging can be determined by the
$k_{tis}$ multiplier (**Figure 5F**) multiplied by the cluster modifier $k_{cl}$
(**Figure 6**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude.

@@

```{r child='../code/figure9.Rmd'}
```

@@

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**Figure 6**). A
cluster-specific rate of 1 represents the typical rate for this tissue type. For
several cancers (_e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held
true across cancer types, which could result from different cancers having
different standards of care for example. In support of this interpretation, we
also tested a factored model which considered both cancer type and each cluster
independent of cancer type (not shown). Though we were able to successfully fit
this model, it is a special case of the more general model where cluster and
tissue independently affect longevity, and there is no reason to believe that
cluster specific effects would necessarily remain constant across tissue types
given how widely the patients vary across tissue types, drug and surgical
treatments.

>>>>

To illustrate the model, we simulate survival time curves
for four related groups (**figure 5A**. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

#### Figure 5
```{r child='../code/survival/figure5.Rmd'}
```

@@

We found that cancer types, as expected, have a range of prognoses relative to
the general population. In **figure 5B-E** we see three particularly deadly
cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
where diagnosis apparently decreases risk relative to background (Thyroid:
THCA). Cancers with posterior probability for relative risk of less than 1 such
as THCA should be interpreted carefully. This Bayesian model is a model for a
state of information. The information that a person is diagnosed with cancer may
lead us to expect that they will live a shorter time than the general population
of matched age (k > 1), or a longer time than the general population of matched
age (k < 1). A shorter prognosis could result from the cancer being aggressive,
injuring the body and causing death. By contrast, a longer prognosis may result
from the cancer being relatively mild, and therefore diagnosis could be an
informational signal that the patient is health conscious, with the comparison
group having more people whose cancers go undiagnosed. It is important to note
therefore that the diagnosis can increase our expectation of life relative to
the comparison group, even if it decreases the expectation of life of the
individual relative to the counterfactual where they did not have cancer.

The degree to which a given cancer accelerates aging can be determined by the
$k_{tis}$ multiplier (**figure 5F**) multiplied by the cluster modifier $k_{cl}$
(**figure 6**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is influenced by the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude.

@@

```{r child='../code/figure6.Rmd'}
```

@@

Our estimates of tissue-specific cluster effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some clusters are either less deadly
or more deadly than other clusters within each cancer type (**figure 6**). A
cluster-specific rate of 1 represents the typical rate for this tissue type. For
several cancers (_e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indistinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples, period. We did not observe cluster-specific trends that held
true across cancer types, which could result from different cancers having
different standards of care for example. In support of this interpretation, we
also tested a factored model which considered both cancer type and each cluster
independent of cancer type (not shown). Though we were able to successfully fit
this model, it is a special case of the more general model where cluster and
tissue independently affect longevity, and there is no reason to believe that
cluster specific effects would necessarily remain constant across tissue types
given how widely the patients vary across tissue types, drug and surgical
treatments.
